Efficient toolkit implementing best practices for principal component analysis of population genetic data

### 2019-12-19 paper-bedpca.tex
merge some suggestions

When applying our automatic procedure to remove long-range LD regions, it does not converge after 5 iterations for the UK Biobank, meaning that it keeps detecting long-range LD regions at each iteration (represented by peaks in PC loadings).
Therefore, we are able to capture only 16 PCs that show stratification that is not LD structure (Figures \ref{fig:UKBB-screeplot}, \ref{fig:UKBB-scores} and \ref{fig:UKBB-loadings}).
Similarly, PC loadings reported by the UK Biobank clearly show that PC19 to PC40 capture LD structure, which is also the case for PC16 and PC18, although less pronounced (see peaks in figure \ref{fig:UKBB-loadings40}). These include [TODO: FIND REGION -> CLOSE TO HLA?] in PC19.

@@

\begin{figure}[htbp]
\centering
\begin{subfigure}[b]{0.5\textwidth}
\includegraphics[width=\textwidth]{hist-outliers-1000G.pdf}
\caption{Distribution of statistics}
\end{subfigure}
\\~\\
\begin{subfigure}[b]{0.45\textwidth}
\includegraphics[width=\textwidth]{outliers-1000G.pdf}
\caption{Principal Component (PC) scores 13 to 20 of 1000G, colored by the statistic (S) used to define outliers.
A few points with higher values for this statistic S appear as outliers in PC17-20.}
\end{subfigure}
~~~~~~
\begin{subfigure}[b]{0.45\textwidth}
\includegraphics[width=\textwidth]{be-outlier-1000G.pdf}
\caption{Principal Component (PC) scores 13 to 20 of 1000G, colored by being detected as an outlier. Threshold of being an outlier is determined based on histogram (a).}
\end{subfigure}
\caption{Outlier detection in the 1000 Genomes (1000G) project, using \texttt{prob\_dist} (Section \ref{outlier-sample}).\label{fig:outlier-pd}}
\end{figure}

@@

In this work, we have compiled different pitfalls that can arise with analyses related to Principal Component Analysis (PCA) of genetic data. Then, we have investigated possible solutions to these pitfalls and chose the ones that we found most relevant, both with respect to properties such as accuracy and robustness, but also that would be fast and easy to use. We then implemented these solutions in open-source R packages. 
The new functions we provide can be directly applied to genotypes stored as PLINK bed/bim/fam files with some missing values.
This contrasts with previous releases of R package bigsnpr that could only use format ``bigSNP''. This data format can store both genotype calls and dosages, but requires conversion from other formats and imputation of missing values using functions provided in the package \cite[]{prive2017efficient}.
As PCA is a useful tool on its own and does not require extensive imputed data, we therefore decided that operating directly on PLINK files with a few missing values would be more practical for users.

As a general recommendation for computing PCA, we propose the pipeline shown in figure \ref{fig:pipeline}. Note that we have not included standard steps such as initial quality control filters and post-analysis checks (e.g.\ visual inspection of different plots).
This pipeline requires removing all related individuals, for which we provide an R wrapper to PLINK's implementation of KING robust kinship coefficients \cite[]{manichaikul2010robust,chang2015second}. 
Note that one should remove both individuals in each related pairs so that the projected individuals are not related to the ones used for computing PCA, since we showed that relatedness is a problem when using the OADP projection (Figure \ref{fig:proj1000G-4}).
After selecting a subset of individuals, we apply several steps of outlier detection, one for outlier variants that capture long-range LD variation (automatic), and one for detecting outlier samples (semi-automatic and visual).
To make these steps more computationally efficient, we explored solutions for not recomputing PCA from scratch when removing a few samples or variants. Using educated guesses in R package PRIMME based on low-rank approximations seemed to be a promising approach but did not reduce computation time by much, so we did not pursue this idea \cite[]{brand2003fast,wu2017primme}. 
Once PCA is done, one should look at the loadings to make sure that there is no visible LD structure left (peaks in PC loadings as shown in figures \ref{fig:UKBB-loadings40} and \ref{fig:UKBB-loadings}), since including PCs that capture LD as covariates in genetic analyses can lead to reduced power for variants in these regions. 
These long-range LD regions are captured by PCs because variants in these regions are extensively correlated (even after a first step of LD thinning), and PCA just captures the variance in these correlated variants. 
This is a similar problem as including related individuals or even whole families in PCA, they would show up as structure in PCs. 
One should also look at PC scores to identify which PCs capture population structure.
As in many applications, we believe there should be a compromise between signal and noise. Therefore, we recommend using only PCs that show structure (e.g.\ PC1-16 in figure \ref{fig:UKBB-scores}) and excluding PCs that indicate LD (e.g.\ PC19-40 in figure \ref{fig:UKBB-loadings40}).

We also provide some recommendations when the dataset is composed mainly of one population (e.g.\ British people in the UK Biobank).
We show that it is useful to actually subset these individuals so there is not too much imbalance between the sample sizes of the different populations. Previous work has also shown that uneven population sizes can distort PCs \cite[]{novembre2008interpreting,mcvean2009genealogical}. 
Indeed, when subsetting British and Irish people, we are able to capture a lot more PCs that show population structure with less than 50K individuals as compared to when using more than 400K individuals who are mostly composed of British and Irish people.
This makes designs such as the 1000 Genomes project, which gathered around 100 people for each of 26 populations, highly relevant for capturing population structure \cite[]{10002015global}.
In contrast, when the goal is to have an homogeneous sample, we show that using the Mahalanobis distance on PC scores can efficiently achieve this goal, which we used in previous analyses \cite[]{prive2019efficient}.
When the homogeneous sample is not predominant in the dataset, one solution is to compute the center and covariance of the robust Mahalanobis distance using only the population of interest, and then computing the distances for all individuals using these robust estimates.

@@

Finally, although we have focused on PCA of genotype data in this paper, we believe most of the results presented here are not inherent to genotype data, and can be transferred to e.g.\ other omics data as well. For example, PCs can be used to account for confounding in other data as well \cite[]{pickrell2010understanding}. Then, outlier and homogeneous sample detection can be used on PCs of other types of data. Moreover, projection of scores will also be a problem for other omics data where the number of variables used is larger than the number of samples used for computing PCA. Finally, using ``populations'' with approximately the same size is relevant for other biological data as well.
However, other pitfalls might apply when using other types of data; for example, methylation data can be confounded by factors such as age and sex, and it might be benificial to remove some methylation probes that are associated with these confounding factors before computing PCA \cite[]{decamps2019guidelines}.

@@

\section*{Declaraction of Interests}

@@

\begin{figure}[htbp]
\centering
\begin{subfigure}[b]{0.5\textwidth}
\includegraphics[width=\textwidth]{hist-outliers2-1000G.pdf}
\caption{Distribution of statistics and default threshold (6, in red)}
\end{subfigure}
\\~\\
\begin{subfigure}[b]{0.45\textwidth}
\includegraphics[width=\textwidth]{outliers2-1000G.pdf}
\caption{Principal Component (PC) scores 13 to 20 of 1000G, colored by statistic used to detect outliers (maximum number of SDs from the mean, log-scale).}
\end{subfigure}
~~~~~~
\begin{subfigure}[b]{0.45\textwidth}
\includegraphics[width=\textwidth]{be-outlier2-1000G.pdf}
\caption{Principal Component (PC) scores 13 to 20 of 1000G, colored by being detected as an outlier.\\}
\end{subfigure}
\caption{Outlier detection in the 1000 Genomes (1000G) project, using the rule ``6 SDs from the mean'', i.e.\ where S2 is the maximum (for all PCs) number of SDs from the mean (Section \ref{outlier-sample}).\label{fig:outlier-sd}}
\end{figure}

>>>>

When applying our automatic procedure to remove long-range LD regions, it does not converge after 5 iterations for the UK Biobank, meaning that it keeps detecting long-range LD regions at each iteration (represented by peaks in PC loadings).
Therefore, we are able to capture only 16 PCs that show stratification that is not LD structure (Figures \ref{fig:UKBB-screeplot}, \ref{fig:UKBB-scores} and \ref{fig:UKBB-loadings}).
Similarly, PC loadings reported by the UK Biobank clearly show that PC19 to PC40 capture LD structure, which is also the case for PC16 and PC18, although less pronounced (see peaks in figure \ref{fig:UKBB-loadings40}). 
These include e.g.\ one region on chromosome 6 (70-91 Mbp) that is captured in PC19 (Figure \ref{fig:UKBB-loadings}) and that was not previously reported in \cite{price2008long}.

@@

\begin{figure}[htbp]
\centering
\begin{subfigure}[b]{0.55\textwidth}
\includegraphics[width=\textwidth]{hist-outliers-1000G.pdf}
\caption{Distribution of statistics (S)}
\end{subfigure}
\\~\\
\begin{subfigure}[b]{0.45\textwidth}
\includegraphics[width=\textwidth]{outliers-1000G.pdf}
\caption{Principal Component (PC) scores 13 to 20 of 1000G, colored by the statistic (S) used to define outliers.
A few points with higher values for this statistic S appear as outliers in PC17-20.}
\end{subfigure}
~~~~~~
\begin{subfigure}[b]{0.45\textwidth}
\includegraphics[width=\textwidth]{be-outlier-1000G.pdf}
\caption{Principal Component (PC) scores 13 to 20 of 1000G, colored by being detected as an outlier. Threshold of being an outlier is determined based on histogram (a).}
\end{subfigure}
\caption{Outlier detection in the 1000 Genomes (1000G) project, using \texttt{prob\_dist} (Section \ref{outlier-sample}).\label{fig:outlier-pd}}
\end{figure}

@@

In this work, we have compiled different pitfalls that can arise with Principal Component Analysis (PCA) of genetic data. Then, we have investigated possible solutions to these pitfalls and selected the ones that we found most advantageous, both with respect to properties such as accuracy and robustness, but also computational efficiency and ease of use. 
We then implemented these solutions in R packages bigsnpr and bigutilsr. 
The new functions we provide in R package bigsnpr can be directly applied to genotypes stored as PLINK bed/bim/fam files with some missing values.
This contrasts with previous releases of bigsnpr that could only use format ``bigSNP''. This data format can store both genotype calls and dosages, but requires conversion from other formats and imputation of missing values using functions provided in the package \cite[]{prive2017efficient}.
As PCA is a useful tool on its own and does not require extensive imputed data, we therefore decided that operating directly on PLINK files with a few missing values would be more practical for users.

We summarise our work into several recommendations for computing PCA, and propose the pipeline shown in figure \ref{fig:pipeline}. Note that we have not included standard steps such as initial quality control filters and post-analysis checks (e.g.\ visual inspection of different plots).
This pipeline requires removing all related individuals, for which we provide an R wrapper to PLINK's implementation of KING robust kinship coefficients \cite[]{manichaikul2010robust,chang2015second}. 
Note that one should remove both individuals in each pair of related individuals. This ensures that the projected individuals are not related to the ones used for computing PCA, since we showed that relatedness is a problem when using the OADP projection (Figure \ref{fig:proj1000G-4}).
After selecting a subset of individuals, we apply several steps of outlier detection, one for outlier variants that capture long-range LD variation (automatic), and one for detecting outlier samples (semi-automatic and visual).
To make these steps more computationally efficient, we explored solutions for not recomputing PCA from scratch when removing a few samples or variants. Using educated guesses in R package PRIMME based on low-rank approximations seemed to be a promising approach but did not reduce computation time by much, so we did not pursue this idea \cite[]{brand2003fast,wu2017primme}. 
Once PCA is done, one should look at the loadings to make sure that there is no visible LD structure left (peaks in PC loadings as shown in figures \ref{fig:UKBB-loadings40} and \ref{fig:UKBB-loadings}), since including PCs that capture LD as covariates in genetic analyses can lead to reduced power for variants in these regions. 
These long-range LD regions are captured by PCs because variants in these regions are extensively correlated (even after a first step of LD thinning), and PCA just captures the variance in these correlated variants. 
This is a similar problem as including related individuals or even whole families in PCA, they would show up as structure in PCs. 
One should also look at PC scores to identify which PCs capture population structure.
As in many applications, we believe there should be a compromise between signal and noise. Therefore, we recommend using only PCs that show structure (e.g.\ PC1-16 in figure \ref{fig:UKBB-scores}) and excluding PCs that indicate LD (e.g.\ PC19-40 in figure \ref{fig:UKBB-loadings40}).

When analysing a dataset that is composed mainly of one population (e.g.\ British people in the UK Biobank), we found that it is useful to subset these individuals to reduce the imbalance between the different population sizes.
Previous work has also shown that uneven population sizes can distort PCs \cite[]{novembre2008interpreting,mcvean2009genealogical}. 
Indeed, when subsetting British and Irish people, we are able to capture a lot more PCs that show population structure with less than 50K individuals as compared to when using more than 400K individuals who are mostly composed of British and Irish people.
This suggests that designs such as the 1000 Genomes project, which gathered around 100 people for each of 26 different populations, are highly relevant for capturing population structure \cite[]{10002015global}.

In contrast, a common strategy in genetic analyses is to restrict the analysis to a homogeneous sample to reduce risk of confounding due to population stratification.
For that purpose, we show that using the Mahalanobis distance on PC scores can efficiently achieve this goal, which we used in previous analyses \cite[]{prive2019efficient}.
When the homogeneous sample is not predominant in the dataset, one solution is to compute the center and covariance of the robust Mahalanobis distance using only the population of interest, and then computing the distances for all individuals using these robust estimates.

We tackled many problems related to computing PCA on genotype data in this paper. 
Yet, more work is still needed. 
For example, the effect of including long-range LD regions in GWAS should be investigated quantitatively.
These are directions of future work.

@@

Finally, although we have focused on PCA of genotype data in this paper, we believe most of the results presented here are not inherent to genotype data, and can be transferred to e.g.\ other omics data as well. For example, PCs can be used to account for confounding in other data as well \cite[]{pickrell2010understanding}. Then, outlier and homogeneous sample detection can be used on PCs of other types of data. Moreover, projection of scores will also be a problem for other omics data where the number of variables used is larger than the number of samples used for computing PCA. Finally, using ``populations'' with approximately the same size is relevant for other biological data as well.
However, other pitfalls might apply when using other types of data; for example, methylation data can be confounded by factors such as age and sex, and it might be beneficial to remove some methylation probes that are associated with these confounding factors before computing PCA \cite[]{decamps2019guidelines}.

@@

\section*{Declaration of Interests}

@@

\begin{figure}[htbp]
\centering
\begin{subfigure}[b]{0.55\textwidth}
\includegraphics[width=\textwidth]{hist-outliers2-1000G.pdf}
\caption{Distribution of statistics (S2) and default threshold (6, in red)}
\end{subfigure}
\\~\\
\begin{subfigure}[b]{0.45\textwidth}
\includegraphics[width=\textwidth]{outliers2-1000G.pdf}
\caption{Principal Component (PC) scores 13 to 20 of 1000G, colored by statistic used to detect outliers (maximum number of SDs from the mean, log-scale).}
\end{subfigure}
~~~~~~
\begin{subfigure}[b]{0.45\textwidth}
\includegraphics[width=\textwidth]{be-outlier2-1000G.pdf}
\caption{Principal Component (PC) scores 13 to 20 of 1000G, colored by being detected as an outlier.\\}
\end{subfigure}
\caption{Outlier detection in the 1000 Genomes (1000G) project, using the rule ``6 SDs from the mean'', i.e.\ where S2 is the maximum (for all PCs) number of SDs from the mean (Section \ref{outlier-sample}).\label{fig:outlier-sd}}
\end{figure}
